A Tool for Range Analysis of Whole Programs

### 2012-05-08 doc/Manuscript/DTool/paper.tex
Fixed some English errors. Improved cohesion by adding some connectives between sentences. Fixed commands syntax in last section.

% Tool capabilities.
We have implemented a range analysis tool on top of the LLVM
compiler~\cite{Lattner04}.
This tool can be used either as a standalone analysis, or it can be called by a
client pass in order to feed this client with more information about the
program.
One of our main concerns when developing our range analysis was scalability,
and we believe that our final product meets the requirements of
industrial-quality software.
We have used it to analyze a test suite with 2.72 million lines of C code.
Our implementation is fast: it globally analyzes the gcc source code in less than
15 seconds, for instance. 
It is also precise: our results are similar to Stephenson
{\em et al}'s~\cite{Stephenson00}, even though our analysis does not require
a backward propagation phase.
Furthermore, we have been able to find tight bounds to the majority of the
examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
In Section~\ref{sec:dsc} we provide a description of our implementation.

% The demo tool.
\noindent
\textbf{Software: }
Our tool is publicly available for
download at \texttt{http://code.google.com/p/range-analysis/}.
In our webpage we provide, in addition to the static range analysis itself,
a profiler that records the minimum and maximum values assigned to each
variable during program execution.
We also provide a visual interval to our tool, which lets the user to
see the control flow graph of the target program.
This visual interface also lets the user to see the integer intervals that
the analysis estimates to each variable.

@@

\noindent
\textbf{A Holistic View of our Range Analysis Algorithm.}
We perform range analysis in a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.
The last phase happens per strong component; however, if we opted for not building 
these components, then it happens once for the entire constraint graph.
Nevertheless, the use of strongly connected components
is so essential for performance and precision that it is considered optional only because we
can easily build our implementation without this module.

@@

Our analysis can be used independently from other compiler clients, to identify
logical problems in the analyzed code.
However, the Range Analysis is more often used as a tool to identify dead code,
memory overflow and redundant checks; hence, enabling other compiler analysis.
We will illustrate this last mode via an example.
In order to use our range analysis, one can write a LLVM pass that calls it.
There is vast documentation~\footnote{\texttt{http://llvm.org/docs/WritingAnLLVMPass.html}} about how to write LLVM passes in the web.
The program below, which is self-contained, is an example of such a pass.
This program is a LLVM pass; that is, it is called as part of the LLVM
tool chain, and receives as input a data-structure of the \texttt{Function}
type, that describes the program.
An iterator over such a data-structure returns a list of basic blocks.
Iterators over basic-blocks return a list of instructions.

%Example pass: Source code
\begin{lstlisting}[frame=single]
//We are ommiting the LLVM includes
#include "../RangeAnalysis/RangeAnalysis.h"
using namespace llvm;
class ClientRA: public llvm::FunctionPass {
public:
  static char ID;
  ClientRA() : FunctionPass(ID){ }
  virtual ~ClientRA() { }
  virtual bool runOnFunction(Function &F){
    IntraProceduralRA<Cousot> &ra = getAnalysis<IntraProceduralRA<Cousot>>();
    errs() << "\nCousot Intra Procedural analysis 
               (Values -> Ranges) of " << F.getName() << ":\n";
    for(Function::iterator bb=F.begin(), bbEnd=F.end(); bb!=bbEnd; ++bb){
      for(BasicBlock::iterator I=bb->begin(), IEnd=bb->end(); I!=IEnd; ++I){
        if(I->getOpcode() == Instruction::Store){
          const Value *v = &(*I);
          Range r = ra.getRange(v);
          r.print(errs());
          I->dump();
        }
      }
    }
    return false;
  }

@@

%Steps to compile and run
In order to use the example client, you need to give it a bitcode input file. Below we show how to do this.
First, we can translate a c file into a bitcode file using clang:
\begin{lstlisting}[frame=single]
clang -c -emit-llvm test.c -o test.bc
\end{lstlisting}
Next thing: we must convert the bitcode file to e-SSA form.
We do it using the \textit{vssa} pass.
The e-SSA conversion module is distributed together with our analysis; however,
its use is not mandatory.
It increases the precision of the analysis, at the expenses of a small slowdown.
In our experiments, we have observed that this slowdown is very small.
\begin{lstlisting}[frame=single]
opt -instnamer -mem2reg -break-crit-edges test.bc -o test.bc
opt -load LLVM_SRC_PATH/Debug/lib/vSSA.so -vssa test.bc -o test.essa.bc
\end{lstlisting}

Notice that we use a number of other passes too, to improve the quality of the
code that we are producing:
\textit{instnamer} just assigns strings to each variable.
We only use this pass for aesthetic reasons.
This pass will help our visual interface to look nicer.
\textit{mem2reg} maps variables allocated in the stack to virtual registers.
Without this pass everything is mapped into memory, and then our range analysis
will not be able to find any meaningful ranges to the variables.
\textit{break-crit-edges} removes the critical edges from the control flow graph
of the input program.
This pass will increase the precision of our range analysis (just a tiny bit
though), because the e-SSA transformation will be able to convert more variables
into a better format.
We run the example client with the commands below:
\begin{lstlisting}[frame=single]
opt -load LLVM_SRC_PATH/Debug/lib/RangeAnalysis.so 
    -load LLVM_SRC_PATH/Debug/lib/ClientRA.so -client-ra test.essa.bc
\end{lstlisting}
This sequence of calls will cause our example pass to be dynamically loaded
by the LLVM framework.
It will receive the bitcodes that we had produced before, and will produce,
as output, a list of variables, followed by their intervals.

>>>>

% Tool capabilities.
We have implemented a range analysis tool on top of the LLVM
compiler~\cite{Lattner04}.
This tool can be used either as a standalone analysis, or it can be called by a
client pass in order to feed this client with more information about the
program.
One of our main concerns when developing our range analysis was scalability,
and we believe that our final product meets the requirements of
industrial-quality software.
We have used it to analyze a test suite with 2.72 million lines of C code.
Our implementation is fast: it globally analyzes the GCC source code in less than
15 seconds, for instance. 
It is also precise: our results are similar to Stephenson
{\em et al}'s~\cite{Stephenson00}, even though our analysis does not require
a backward propagation phase.
Furthermore, we have been able to find tight bounds to the majority of the
examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
In Section~\ref{sec:dsc} we provide a description of our implementation.

% The demo tool.
\noindent
\textbf{Software: }
Our tool is publicly available for
download at \texttt{http://code.google.com/p/range-analysis/}.
In our webpage we provide, in addition to the static range analysis itself,
a profiler that records the minimum and maximum values assigned to each
variable during program execution.
We also provide a visual interface to our tool, which lets the user to
see the control flow graph of the target program.
This visual interface also lets the user to see the integer intervals that
the analysis estimates to each variable.

@@

\noindent
\textbf{A Holistic View of our Range Analysis Algorithm.}
We perform range analysis in a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.
The last phase happens per strong component; however, if we opted not for building 
these components, then it would happen once for the entire constraint graph.
Nevertheless, the use of strongly connected components
is so essential for performance and precision that it is considered optional only because we
can easily build our implementation without this module.

@@

Our analysis can be used independently from other compiler clients to identify
logical problems in the analyzed code.
However, the Range Analysis is more often used as a tool to identify dead code,
memory overflow and redundant checks; hence, enabling other compiler analyses.
We will illustrate this last mode via an example.
In order to use our range analysis, one can write a LLVM pass that calls it.
For this purpose, there is vast documentation~\footnote{\texttt{http://llvm.org/docs/WritingAnLLVMPass.html}} about how to write LLVM passes.
This program below is an example of LLVM client pass, that is, it is called as part of the LLVM
tool chain, and receives as input a data-structure of the \texttt{Function}
type, that describes the program.
An iterator over such a data-structure returns a list of basic blocks.
Consequently, iterators over basic-blocks return a list of instructions.

%Example pass: Source code
\begin{lstlisting}[frame=single]
//We are omitting the LLVM includes
#include "../RangeAnalysis/RangeAnalysis.h"
using namespace llvm;
class ClientRA: public llvm::FunctionPass {
public:
  static char ID;
  ClientRA() : FunctionPass(ID){ }
  virtual ~ClientRA() { }
  virtual bool runOnFunction(Function &F){
    IntraProceduralRA<Cousot> &ra = getAnalysis<IntraProceduralRA<Cousot>>();
    errs() << "\nCousot Intra Procedural analysis 
               (Values -> Ranges) of " << F.getName() << ":\n";
    for(Function::iterator bb=F.begin(), bbEnd=F.end(); bb!=bbEnd; ++bb){
      for(BasicBlock::iterator I=bb->begin(), IEnd=bb->end(); I!=IEnd; ++I){
        if(I->getOpcode() == Instruction::Store){
          const Value *v = &(*I);
          Range r = ra.getRange(v);
          r.print(errs());
          I->dump();
        }
      }
    }
    return false;
  }

@@

%Steps to compile and run
In order to use the example client, you need to give it a bitcode input file. Below we show how to do this.
Firstly, we must translate a c file into a bitcode file using clang:
\begin{lstlisting}[frame=single]
clang -c -emit-llvm test.c -o test.bc
\end{lstlisting}

Next step: we must convert the bitcode file to e-SSA form.
We do it using the \textit{vssa} pass.
The e-SSA conversion module is distributed together with our analysis; however,
its use is not mandatory.
It increases the precision of the analysis, at the expense of a small slowdown.
\begin{lstlisting}[frame=single]
opt -mem2reg -instnamer -break-crit-edges test.bc -o test.bc
opt -load LLVM_SRC_PATH/BUILD_TYPE/lib/vSSA.so -vssa test.bc -o test.essa.bc
\end{lstlisting}

Notice that we use a number of other passes too, to improve the quality of the
code that we are producing:
\textit{instnamer} just assigns strings to each variable.
We only use this pass for aesthetic reasons, as it 
will help our visual interface to look nicer.
\textit{mem2reg} maps variables allocated in the stack to virtual registers.
Without this pass everything is mapped into memory, and then our range analysis
will not be able to find any meaningful ranges to the variables.
\textit{break-crit-edges} removes the critical edges from the control flow graph
of the input program.
Therefore, this pass will increase the precision of our range analysis (just a tiny bit
though), because the e-SSA transformation will be able to convert more variables
into a better format.
We run the example client with the commands below:
\begin{lstlisting}[frame=single]
opt -load LLVM_SRC_PATH/BUILD_TYPE/lib/RangeAnalysis.so 
    -load LLVM_SRC_PATH/BUILD_TYPE/lib/ClientRA.so -client-ra test.essa.bc
\end{lstlisting}

This sequence of calls will cause our example pass to be dynamically loaded
by the LLVM framework.
It will receive the bitcodes that we had produced before, and will produce,
as output, a list of variables, followed by their intervals.
